Mixing Time Estimation in Reversible Markov Chains from a Single Sample Path

### 2015-08-03 nips2015/rebuttal.txt

We thank the referees for the helpful comments. A detailed response follows.

Tightness of bounds: We don't claim tightness of the bounds -- they
are indeed rather loose -- but our results give the first non-trivial
(upper and lower) bounds in this challenging setting. Our work are
motivated by practical applications where fully data-dependent
intervals were sought but not actually achieved. Our results are a
first step towards a practical solution: we demonstrate feasibility
and also prove limitations on what is possible.

Relaxation vs mixing time: In the revision, we will be more explicit
about estimating the relaxation time rather than mixing time proper
(throughout the abstract and paper). We agree that this is an
important distinction that should be emphasized.

Reviewer_1

3. Qualitatively, Theorems 1 and 2 show that if n is less than some
function of (d,pi,gamma), then for any estimator based on a sample
path of length n, there is some Markov chain such that the error is
large.

4. If n > max{16C^2,4C}/(pi_* gamma_*), then the RHS in (4) (without
the logarithmic factor) is less than pi_*/2.

If n > max{16C^2,4C}/(pi_* gamma_*^3), then the RHS in (5) (without
the logarithmic factor) is less than gamma_*/2.

5. Combining Theorems 1 and 2 with a standard argument yields the
lower bound in the introduction. We will include the details in the
revision.

6.

7. To conclude Theorem 2, it follows from Paley-Zygmund and the moment
bounds that, with probability > 1/4, the earliest time in which a
sample path visits all d states is at least c d log(d) / \bar\gamma.
This implies that an estimator using a shorter sample path cannot
distinguish P^0 from P^i for some i (with probability at least 1/4),
and hence must have large estimation error.

Reviewer_2

\tilde{O}: We'll use this more standard use in the revision.

Theorem 2: Indeed, including an extra constant factor in the claim (or
minding inequality strictness) resolves this issue.

Lemma 2: We'll include more details in the revision.

Theorem 2 (line 958): The estimator cannot rule out P^i without having
vistited state i. We'll clarify this point in the revision.

Reviewer_7

Perfect sampling: Indeed, this line of work is related, although these
methods generally do not apply in our setting of estimation from a
single sample path. We'll discuss this in the revision.

### 2015-08-03 nips2015/rebuttal.txt

We thank the referees for the helpful comments. We'll incorporate the
suggestions in a revision of the paper, especially where more detailed
explanations are needed.

The following two concerns were common to many of the reviews.

@@

4. If n > max{16C^2,4C}/(pi_* gamma_*), then the RHS in (4) (without
the logarithmic factor) is less than pi_*/2.

If n > max{16C^2,4C}/(pi_* gamma_*^3), then the RHS in (5) (without
the logarithmic factor) is less than gamma_*/2.

5. Combining Theorems 1 and 2 with a standard argument yields the
lower bound in the introduction. We will include the details in the
revision.

6. ???

7. To conclude Theorem 2, it follows from Paley-Zygmund and the moment
bounds that, with probability > 1/4, the earliest time in which a
sample path visits all d states is at least c d log(d) / \bar\gamma.
This implies that an estimator using a shorter sample path cannot
distinguish P^0 from P^i for some i (with probability > 1/4), and
hence must have large estimation error. We'll include these details in
the revision.

### 2015-08-05 nips2015/rebuttal.txt

We thank the referees for the helpful comments. We'll incorporate the
suggestions in a revision of the paper, especially where more detailed
explanations are needed.

The following two concerns were common to many of the reviews.

Tightness of bounds: We don't claim tightness of the bounds -- they
are indeed rather loose -- but our results give the first non-trivial
(upper and lower) bounds in this challenging setting. Our work are
motivated by practical applications where fully data-dependent
intervals were sought but not actually achieved. Our results are a
first step towards a practical solution: we demonstrate feasibility
and also prove limitations on what is possible.

Relaxation vs mixing time: In the revision, we will be more explicit
about estimating the relaxation time rather than mixing time proper
(throughout the abstract and paper). We fully agree that this is a
very important distinction that should be emphasized, and we'll amend
the overstated claims throughout.

@@

We will add these details in the revision.

5. Combining Theorems 1 and 2 with a standard argument yields the
lower bound in the introduction. We will include the details in the
revision.

6. ???

7. To conclude Theorem 2, it follows the second-moment method (e.g.,
Paley-Zygmund) and the moment bounds that, with probability > 1/4, the
earliest time in which a sample path visits all d states is at least c
d log(d) / \bar\gamma.  This implies that an estimator using a shorter
sample path cannot distinguish P^0 from P^i for some i (with
probability > 1/4), and hence must have large estimation error. We'll
include these details in the revision.

@@

Reviewer_2

\tilde{O}: We'll adopt this more standard use in the revision.

### 2015-08-07 nips2015/rebuttal.txt

We thank the referees for the helpful comments. We'll incorporate the
suggestions in a revision of the paper, especially where more detailed
explanations are needed.

The following two concerns were common to multiple reviews.

Tightness of bounds: We don't claim tightness of the bounds -- they
are indeed rather loose -- but our results give the first non-trivial
(upper and lower) bounds in this challenging setting. Our work is
motivated by practical applications where fully data-dependent
intervals were sought but not actually achieved. Our results are a
first step towards a practical solution: we demonstrate feasibility
and also prove limitations on what is possible.

@@

Relaxation vs mixing time: In the revision, we will be more explicit
about estimating the relaxation time rather than mixing time proper
(throughout the abstract and paper). We fully agree that this is a
very important distinction that should be emphasized, and we'll amend
the overstated claims throughout.

Additional responses to individual reviews follow.

Reviewer_1

2. We'll make this edit in the revision.

3. Qualitatively, Theorems 1 and 2 show that if n is less than some
function of (d,pi,gamma), then for any estimator based on a sample
path of length n, there is some Markov chain such that the error is
large. We'll add additional clarifying interpretations in the
revision.

@@

If n > max{16C^2,4C}/(pi_* gamma_*^3), then the RHS in (5) (without
the logarithmic factor) is less than gamma_*/2.

We will add these details in the revision.

5. Combining Theorems 1 and 2 with a standard argument yields the
lower bound in the introduction. We will include the details in the
revision.

6. Unfortunately, we were not able to understand the counterexample,
likely due to typos in the review. We hope to resolve this via the
program chairs/area chair.

@@

7. To conclude Theorem 2, it follows the second-moment method (e.g.,
Paley-Zygmund) and the moment bounds that, with probability > 1/4, the
earliest time in which a sample path visits all d states is at least c
d log(d) / \bar\gamma.  This implies that an estimator using a shorter
sample path cannot distinguish P^0 from P^i for some i (with
probability > 1/4), and hence must have large estimation error. We'll
include these details in the revision.

Reviewer_2

Theorem 5: We get an improvement when the min in the bound for
|\hat{V}| is achieved by the first term. A sheepish caveat: this is
not always the case for all sufficiently large n, due to the
sqrt{log(n)} term in the numerator, but will be true for a very large
range of n. There is more discussion in Section 5.

@@

\tilde{O}: We'll adopt this more standard use in the revision.

kappa: It is defined just above Theorem 4.

Theorem 2: Indeed, including an extra constant factor in the claim (or
minding inequality strictness) resolves this issue.

Theorem 2 (line 958): An estimator cannot rule out P^i without having
vistited state i. We'll clarify this point in the revision.

### 2015-08-07 nips2015/rebuttal.txt

If n > max{16C^2,4C}/(pi_* gamma_*^3), then the RHS in (5) (without the logarithmic factor) is less than gamma_*/2.

We will add these details in the revision.

5. Combining Theorems 1 and 2 with a standard argument yields the lower bound in the introduction. We will include the details in the revision.

6. Regarding Theorem 1: Unfortunately, we were not able to understand the counterexample, partly due to typos in the review. We hope to resolve this via the program chairs/area chair.

However, there were some minor numerical calculation errors in the proof of Theorem 1, which we can clarify here. We set \varepsilon := \bar\pi, and observe that \pi^{(2)} = (1/(1+2\varepsilon),2\varepsilon/(1+2\varepsilon)). The rest of the proof is correct as is.

@@

Theorem 5: We get an improvement when the min in the bound for |\hat{V}| is achieved by the first term. A sheepish caveat: this is not always the case for all sufficiently large n, due to the sqrt{log(n)} term in the numerator, but will be true for a very large range of n. There is more discussion in Section 5.

\tilde{O}: We'll adopt this more standard use in the revision.

kappa: It is defined just above Theorem 4.

Theorem 2 (line 188): Indeed, including an extra constant factor in the claim (or minding inequality strictness) resolves this issue.

Theorem 2 (line 958): An estimator cannot rule out P^i without having visited state i. We'll clarify this point in the revision.

### 2015-08-07 nips2015/rebuttal.txt

We will add these details in the revision.

5. Combining Theorems 1 and 2 with a standard argument yields the lower bound in the introduction. We will include the details in the revision.

6. Regarding Theorem 1: Unfortunately, we were not able to understand the counterexample, partly due to typos in the review. We hope to resolve this via the program chairs/area chair.

However, there were some minor numerical calculation errors in the proof of Theorem 1, which we can clarify here. We need to change the "1/6" in the theorem statement and proof to "1/8". Furthermore, in the proof, \bar\pi should be taken in (0,1/4), and we should set \varepsilon := \bar\pi. Observe that \pi^{(2)} = (1/(1+2\varepsilon),2\varepsilon/(1+2\varepsilon)). Thus, \pi_\star \geq \bar\pi in both P^{(1)} and P^{(2)}. To guarantee |\hat\gamma_\star - \gamma_\star| < 1/8, it must be possible to distinguish the two Markov chains. The rest of the proof is correct as is.

@@

Theorem 5: We get an improvement when the min in the bound for |\hat{V}| is achieved by the first term. A sheepish caveat: this is not always the case for all sufficiently large n, due to the sqrt{log(n)} term in the numerator, but will be true for a very large range of n. There is more discussion in Section 5.

\tilde{O}: We'll adopt this more standard use in the revision.

kappa: It is defined just above Theorem 4.

Theorem 2 (line 188): Indeed, including an extra constant factor in the claim resolves this issue.

### 2015-08-07 arxiv-v2/abstract.tex
arxiv-v2 (not yet submitted)

%!TEX root =  matrix-est.tex
This article provides the first procedure for computing a fully
data-dependent interval that traps the mixing time $\tmix$ of a finite
reversible ergodic Markov chain at a prescribed confidence level.
The interval is computed from a single finite-length sample
path from the Markov chain, and does not require the 
knowledge of any parameters of the chain.
%bounds on the mixing time itself.
This stands in contrast to previous approaches, which either 
only provide point estimates, or require a
reset mechanism, or additional prior knowledge.
%Most notably, 
%the procedure produces a fully empirical confidence
%interval (i.e., it depends only on the sample path and a given
%confidence parameter $\delta$) that is always valid with probability
%$1-\delta$; and 
The width of the interval converges to zero at a
$\sqrt{n}$ rate, where $n$ is the length of the sample path. \todoc{Aryeh prefers: .. at a rate of $\sqrt{n}$. }
Upper and lower bounds are given on the number of samples required to
achieve constant-factor multiplicative accuracy.
\todoc{We may need to add some explanation to connect $\gap$ estimates
to $\tmix$ estimates.}
The lower bounds indicate that, unless further restrictions are placed
on the chain, no procedure can achieve this accuracy level before
seeing each state at least $\Omega(\tmix)$ times on the average.
Finally, future directions of research are identified.
%If $\gap$ denotes the absolute spectral gap of the unknown Markov chain
%and the chain's stationary distribution is minorized by $\pimin$,
%our procedure requires $\tilde{O}( \frac{1}{\pimin\gap^3})$ observations
% to estimate the mixing time up to a constant multiplicative accuracy, 
%while we also establish that any algorithm to achieve the same
%goal would need $\Omega( d\log(d)/\gap + 1/\pimin)$ samples.

%and the dependence on the main  chain's

%We provide confidence guarantees that are fully data-dependent in the sense that they yield non-trivial
%finite-sample deviation bounds without any knowledge of the mixing coefficients.
%
%
%The novelty of our approach lies in several key aspects.
%Crucially,
%we do not assume any a priori bounds
%on the mixing time,
%which
%sets our approach apart from
%most others that do
%require some advance knowledge of this quantity.
%Another
%advantage of our result over
%common practice is the sufficiency of a single long sampling sequence,
%without access to a restart mechanism.
%Perhaps most importantly, we overcome an obstacle inherent in the chicken-and-egg nature of
%estimating mixing coefficients: to give confidence bounds one typically needs to know the mixing time.
%We provide confidence guarantees that are fully data-dependent in the sense that they yield non-trivial
%finite-sample deviation bounds without any knowledge of the mixing coefficients.
%%Upper and lower sample complexity bounds are provided and challenging open problems are posed.
%%\todod{I would remove the last sentence, at least the part about
%%``challenging'' open problems.}
%%%test

>>>>

This article provides the first procedure for computing a fully data-dependent interval that traps the mixing time $\tmix$ of a finite reversible ergodic Markov chain at a prescribed confidence level.  The interval is computed from a single finite-length sample path from the Markov chain, and does not require the knowledge of any parameters of the chain.  This stands in contrast to previous approaches, which either only provide point estimates, or require a reset mechanism, or additional prior knowledge.  The width of the interval converges to zero at a $\sqrt{n}$ rate, where $n$ is the length of the sample path.  Upper and lower bounds are given on the number of samples required to achieve constant-factor multiplicative accuracy.  The lower bounds indicate that, unless further restrictions are placed on the chain, no procedure can achieve this accuracy level before seeing each state at least $\Omega(\tmix)$ times on the average.  Finally, future directions of research are identified.

### 2015-08-07 arxiv-v2/conclusions.tex
arxiv-v2 (not yet submitted)

The construction used in \cref{thm:combined} applies more generally:
Given a confidence interval of the form $I_n = I_n(\gap,\pimin,\delta)$ 
for some confidence level $\delta$
and a fully empirical confidence set $E_n(\delta)$ for $(\gap,\pimin)$  for the same level,
$I_n' = E_n(\delta) \cap \cup_{(\gamma,\pi)\in E_n(\delta)} I_n(\gamma,\pi,\delta)$ is a valid
fully empirical $2\delta$-level confidence interval whose asymptotic width
matches that of $I_n$ up to lower order terms under reasonable assumptions on $E_n$ and $I_n$.
In particular, this suggests that future work should focus on 
closing the gap between the lower and upper bounds on the accuracy
of point-estimation. Another interesting direction is to 
reduce the computation cost: The current cubic cost in the number of states
can be too high even when the number of states is moderate.

>>>>

The construction used in \cref{thm:combined} applies more generally:
Given a confidence interval of the form $I_n = I_n(\gap,\pimin,\delta)$ 
for some confidence level $\delta$
and a fully empirical confidence set $E_n(\delta)$ for $(\gap,\pimin)$  for the same level,
$I_n' = E_n(\delta) \cap \cup_{(\gamma,\pi)\in E_n(\delta)} I_n(\gamma,\pi,\delta)$ is a valid
fully empirical $2\delta$-level confidence interval whose asymptotic width
matches that of $I_n$ up to lower order terms under reasonable assumptions on $E_n$ and $I_n$.
In particular, this suggests that future work should focus on 
closing the gap between the lower and upper bounds on the accuracy
of point-estimation. Another interesting direction is to 
reduce the computation cost: The current cubic cost in the number of states
can be too high even when the number of states is only moderately
large.

### 2015-08-07 arxiv-v2/lowerbound.tex
arxiv-v2 (not yet submitted)

\begin{proof}[Proof of \cref{thm:lb-pimin}]
  Fix $\bar\pi \in (0,1/3)$, and set $\veps := 2\bar\pi/(1-\bar\pi)$.
  Consider two Markov chains given by the following stochastic
  matrices:
  \[
    \vP^{(1)} :=
    \begin{bmatrix}
      1-\veps & \veps \\
      1-\veps & \veps
    \end{bmatrix}
    , \quad
    \vP^{(2)} :=
    \begin{bmatrix}
      1-\veps & \veps \\
      1/2 & 1/2
    \end{bmatrix}
    .
  \]
  Each Markov chain is ergodic and reversible; their stationary
  distributions are, respectively, $\vpi^{(1)} = (1-\veps,\veps)$
  and $\vpi^{(2)} = (\veps/(2+\veps),2/(2+\veps))$.
  We have $\pimin \geq \bar\pi$ in both cases.
  For the first Markov chain, $\slem = 0$, and hence the spectral gap
  is $1$; for the second Markov chain, $\slem = 1/2-\veps$, so the
  spectral gap is $1/2+\veps$.

  In order to guarantee $|\hatgap - \gap| < 1/6 < 1/2-\veps$, it must
  be possible to distinguish the two Markov chains.
  Assume that the initial state distribution has mass at least $1/2$
  on state $1$.
  (If this is not the case, we swap the roles of states $1$ and $2$ in
  the constructions above.)
  With probability at least half, the initial state is $1$; and both
  chains have the same transition probabilities from state $1$.
  The chains are indistinguishable less the sample path eventually
  reaches state $2$.
  But with probability at least $1/4$, a sample path of length $n <
  c/\bar\pi = c(2+\veps)/\veps$ starting from state $1$ always
  remains in the same state.
\end{proof}

>>>>

\begin{proof}[Proof of \cref{thm:lb-pimin}]
  Fix $\bar\pi \in (0,1/4)$.
  Consider two Markov chains given by the following stochastic
  matrices:
  \[
    \vP^{(1)} :=
    \begin{bmatrix}
      1-\bar\pi & \bar\pi \\
      1-\bar\pi & \bar\pi
    \end{bmatrix}
    , \quad
    \vP^{(2)} :=
    \begin{bmatrix}
      1-\bar\pi & \bar\pi \\
      1/2 & 1/2
    \end{bmatrix}
    .
  \]
  Each Markov chain is ergodic and reversible; their stationary
  distributions are, respectively, $\vpi^{(1)} = (1-\bar\pi,\bar\pi)$
  and $\vpi^{(2)} = (1/(1+2\bar\pi),2\bar\pi/(1+2\bar\pi))$.
  We have $\pimin \geq \bar\pi$ in both cases.
  For the first Markov chain, $\slem = 0$, and hence the spectral gap
  is $1$; for the second Markov chain, $\slem = 1/2-\bar\pi$, so the
  spectral gap is $1/2+\bar\pi$.

  In order to guarantee $|\hatgap - \gap| < 1/8 < |1 -
  (1/2+\bar\pi)|/2$, it must be possible to distinguish the two Markov
  chains.
  Assume that the initial state distribution has mass at least $1/2$
  on state $1$.
  (If this is not the case, we swap the roles of states $1$ and $2$ in
  the constructions above.)
  With probability at least half, the initial state is $1$; and both
  chains have the same transition probabilities from state $1$.
  The chains are indistinguishable less the sample path eventually
  reaches state $2$.
  But with probability at least $1/4$, a sample path of length $n <
  c/\bar\pi$ starting from state $1$ always remains in the same state.
\end{proof}
